Error-Centric Annotation of Learner Corpora

### 2009-04-09 text/paper.tex
Finally?

\documentclass[12pt]{article}
\usepackage[utf8x]{inputenc}
\usepackage{graphicx,linguex,natbib,aleks,amsmath,amsthm}
\bibpunct{(}{)}{;}{a}{,}{,}
\bibliographystyle{plainnat}
\author{Magdalena Leshtanska \and Aleksandar Dimitrov}
\title{Error-Centric Annotation of Learner Corpora}
\begin{document}
\maketitle

@@

We propose a novel approach to annotating learner errors, based on the idea that
an error is an inconsistency with its surroundings. This basic markup approach
can be  combined with a taxonomy for error classification. We
have coupled it with learner error  classification system and used the resulting
scheme to annotate a subset of the NOCE corpus (\cite{negrilloph}), a corpus of
beginning to intermediate English learners of Spanish.

Section \ref{sec:scheme} will introduce the general idea behind our annotation
scheme and elaborate on more technical details. Section 3  gives a brief
overview of real world performance of our method, as far as this is indeed
possible. An annotation manual is contained in Appendix A.

@@

Our annotation scheme presents three distinct kinds of errors:

@@

\begin{itemize}
  \item \textit{Normative Errors.} Some normative concepts are present in almost
  all European languages, such as capitalization at the beginning of a sentence.
  We chose to ignore such errors, since they are not a sign of weakness in the
  target language but likely pure by chance mistakes.
  \item \textit{Style Errors.} Style is too soft a concept. Errors in style are
  notoriously hard to peg and reliably quantify. Moreover, our corpus
  contained data mostly from beginning English learners, where style errors are
  not as important, evident or relevant.
  \item \textit{Punctuation.} Punctuation mistakes are a very delicate concern:
  they can have tremendous effects on any given token string on both a syntactic
  and semantic level. They might change constituent boundaries, sentence
  boundaries, argument structures and many more things. Moreover, they tend to
  have cascading effects on the correctness of a given sentence. Again, because
  we were annotating a beginner's corpus, we chose to ignore these kinds of
  errors, since in beginner language punctuation often degrades to line noise
  very quickly.
\end{itemize}

@@

Based on these assumptions, we decided to decouple the \textit{error markup}
from the \textit{corpus data}. Specifically, to our annotation method, the
errors and the text are two entirely different data structures. Every particular
error can reference tokens within the corpus using a \textbf{key}, similar to
the way modern Relational Data Bases reference their data. This makes the index
more accessible and easier to maintain and eliminates both problems above,
because every single annotation can be completely independent from all other
annotations.  The source files for our corpus consisted of plain text files from
the NOCE-corpus of beginning to intermediate Spanish learners of English. Our
corpus format is defined in \ref{def:errorformat}.

@@

Existing inter-annotator agreement measures all assume the presence of atomic
units in the corpus data, which are annotated by the annotators of a certain
data set.  The annotations \textit{over these units} are then used to calculate an
agreement coefficient. However, constituents of learner language, being fairly
diverse in nature, are not as easily contained in atomic units. Instead, a
common task in the annotation of learner language is \textit{delineating}
the extent of a certain mistake.

\cite{artstein} briefly discuss unitization and go on to note that it has thus far
not been exhaustively researched. Even more importantly, they explicitly comment
on the unknown status of the validity of the only inter-annotator agreement
measure in the corpus linguistic literature, $\alpha_U$, presented in
\cite{krip}.  Apart from being untested, $\alpha_U$ also seems to have
problems with overlapping markup both in the text, and between annotation sets
(both of which are frequent in learner language data.) The measure also assumes
annotation spans to be continuous, which is not the case in our data.

@@

Table \ref{tab:1} shows the total amount of corresponding data in the individual
markup. The table's labeling reads as follows: $s$ stands for a non-empty
intersection between the two data sets (or prefix relation for the error types)
and $i$ for equality.  $o,y,r$ denote the data type: $o$ stands for error
tOkens, $y$ for the error's tYpe, and $r$ for the errors context (or
\textit{tRigger}). Thus, $so\wedge sy\wedge ir$ is the number of all annotations
that have a partial overlap on the error tokens $E$, a prefix match on the error
type $\theta$, and identical error context.

>>>>

\documentclass{article}
\usepackage[utf8x]{inputenc}
\usepackage{graphicx,linguex,natbib,aleks,amsmath,amsthm}
\bibpunct{(}{)}{;}{a}{,}{,}
\bibliographystyle{plainnat}
\author{Magdalena Leshtanska \and Aleksandar Dimitrov}
\title{Error-Centric Annotation of Learner Corpora}
\begin{document}
\maketitle

@@

We propose a novel approach to annotating learner errors, based on the idea that
an error is an inconsistency with its context. This basic markup approach
can be  combined with a taxonomy for error classification. We
have coupled it with learner error  classification system and used the resulting
scheme to annotate a subset of the NOCE corpus (\cite{negrilloph}), a corpus of
beginner to intermediate English learners of Spanish.

Section \ref{sec:scheme} will introduce the general idea behind our annotation
scheme and elaborate on more technical details. Section 3  gives a brief
overview of real world performance of our method, as far as this is indeed
possible. Appendix A contains the annotation manual.

@@

Our annotation scheme presents three distinct kinds of errors based on a
definition of an error as an inconsistency with the context:

@@

\begin{itemize}
  \item \textit{Normative Errors.} Some normative concepts are present in almost
  all European languages, such as capitalization at the beginning of a sentence.
  We chose to ignore such errors, since they are not a sign of weakness in the
  target language but likely pure by chance mistakes or general incompetence.
  \item \textit{Style Errors.} Style is too soft a concept. Errors in style are
  notoriously hard to peg and reliably quantify. Moreover, our corpus
  contained data mostly from beginner English learners, where style errors are
  not as important, evident or relevant.
  \item \textit{Punctuation.} Punctuation mistakes are a very delicate concern:
  they can have tremendous effects on any given token string on both a syntactic
  and semantic level. They might change constituent boundaries, sentence
  boundaries, argument structures and many more things. Moreover, they tend to
  have cascading effects on the correctness of a given sentence. Again, because
  we were annotating a beginner's corpus, we chose to ignore these kinds of
  errors, since in beginner language punctuation often degrades to line noise
  very quickly.
\end{itemize}

@@

Based on these assumptions, we decided to decouple the \textit{error markup}
from the \textit{corpus data}. Specifically, to our annotation method, the
errors and the text are two entirely different data structures. Every particular
error can reference tokens within the corpus using a \textbf{key}, similar to
the way modern Relational Data Bases reference their data. This makes the index
more accessible and easier to maintain and eliminates both problems above,
because every single annotation can be completely independent from all other
annotations. Our corpus format is defined in \ref{def:errorformat}.

@@

Existing inter-annotator agreement measures all assume the presence of atomic
units in the corpus data, which are annotated by the annotators of a certain
data set.  The annotations \textit{over these units} are then used to calculate an
agreement coefficient. However, constituents of learner language, being fairly
diverse in nature, are not as easily contained in atomic units. Instead, a
common task in the annotation of learner language is \textit{delineating}
the extent of a certain mistake. This has been called \textbf{unitization} by
\cite{krip}.

\cite{artstein} briefly discuss unitization and go on to note that it has thus far
not been exhaustively researched. Even more importantly, they explicitly comment
on the unknown status of the validity of the only inter-annotator agreement
measure in the literature, $\alpha_U$, presented in
\cite{krip}.  Apart from being untested, $\alpha_U$ also seems to have
problems with overlapping markup both in the text, and between annotation sets
(both of which are frequent in learner language data.) The measure also assumes
annotation spans to be continuous, which is not the case in our data.

@@

Table \ref{tab:1} shows the total amount of corresponding data in the individual
markup. The table's labeling reads as follows: $s$ stands for a non-empty
intersection between the two data sets' fields. For the error types, $s$ was
defined to be a prefix operator instead, accounting for common paths along the
taxonomy tree.  $i$ stands for equality.  $o,y,r$ denote the data type: $o$
stands for error tOkens, $y$ for the error's tYpe, and $r$ for the errors
context (or \textit{tRigger}). Thus, $so\wedge sy\wedge ir$ is the number of all
annotations that have a partial overlap on the error tokens $E$, a prefix match
on the error type $\theta$, and identical error context.

Having a look at the data presented in table \ref{tab:1}, the figures do not
seem to be in strong support of our efforts thus far, although there is little
grounds for comparison with other similar data. Nevertheless the figures may
indeed allow partial assessment of the data.  Less than half of the annotations
were completely equivalent. Interestingly, while the discrepancy between a
strict and a fuzzy match on the token sequences alone is above 10\%, it is very
low when we demand matches on the rest of the data structure, too. This probably means
that we have an overhead of unclear cases, which are currently hard to catch,
whereas on the other hand we have a clear cut kernel of reliable rules, too.

Upon inspecting the corpus data in greater detail, we found out that it was
sometimes hard to decide whether a particular error was a contextual or
grammatic error. Moreover, often the categories within the grammatic error
branch were not clear enough. It seems that our merger of the CLC and Free Text
introduced some confusion with certain categories. Particularly, the
\textit{omission}, \textit{replace}, and \textit{redundancy} cases would at
times `steal' hits from the other categories due to their very general
nomenclature. We have revised the annotation manual to specifically mention that
all other grammatical error categories take precedence over these three.
Unfortunately, this seems to be a suboptimal solution. The next section contains
a proposal for a much cleaner solution.

In revising the annotation scheme, the comment field of our annotations proved
to be quite useful. Both annotators commented frequently that they lacked
categories for adverbs and gerunds. The former has been integrated in to the
scheme already, and we might add a gerund field in the future, too.

Another very unclear point in the original manual was the minimalization of
error spans constraint, which advises the annotator to only select the smallest
possible amount of tokens, even when that means splitting up compound nouns.
Also, both annotators reacted differently to situations where a sequence had to
be marked up. This spawned a new subsection in our annotation manual as well.
